---
code-annotations: hover
---

```{r}
#| label: setup
#| include: false

pacman::p_load(tidyverse, palmerpenguins)
```

# Annotations and titles

Besides the layers required to generate a visualisation, additional layers can be added enhance the messages given by the data, drawing readers’ attention to interesting findings and the story you are trying to tell. 

## Plot and axis titles
In the previous section, we saw how axis and legend titles can be added within `scale` functions using the `name` argument. `ggplot2` also contains the `labs` function which can be added as a separate layer to control these titles and can also add plot titles, subtitles and footnotes:

```{r}
#| label: labs-example
#| warning: false

ggplot(data = penguins, 
       aes(x = year, fill = species)) +
  geom_bar(colour = "black") +
  scale_fill_brewer(palette = "Dark2") +
  labs(title = "This is the plot title",
      subtitle = "This is a subtitle",
      caption = "This is a caption, useful for source information.",
      x = "x-axis title", y = "y-axis title", 
      fill = "Legend title")  
```

:::{.stylebox}

::::{.stylebox-header}
:::::{.stylebox-icon}
:::::
Style tip
::::

::::{.stylebox-body}
Mathematical equations can be added into `labs` arguments by surrounding the text with the `quote()` function. Check `?plotmath` for examples of equation syntax.

`\n` can be used to specify line breaks within the `labs` arguments.

Specifying any of these arguments as `NULL` (no speech marks) removes the title from the visualisation.
::::

:::

## Annotations
Annotations can be useful to include context to visualisations and draw attention to important messages. Annotations can include text labels, reference lines and shading, amongst others. `ggplot2` contains a number of `geom` objects that can be used to add annotation layers to a visualisation. As these annotations are added within `geom`s, they can be specified using values from the data (when wrapped in the `aes` function) or manually. This section will cover some common annotations but there are many others available (see the [ggplot ebook](https://ggplot2-book.org/annotations) for a more comprehensive list).

### Text labels
Text labels can either be added using `geom_text` or `geom_label` (which adds text surrounded by a rectangular box, making it easier to read on busy backgrounds). Aesthetics such as `x`, `y` and `colour` can be used to customise text labels (either manually or from the data). Other aesthetics that can be added include:

- `label` defines the text displayed
- `angle` rotates the text
- `family` defines the font 
- `fontface` can be changed to make text `"bold"` or `"italic"`


```{r}
#| label: fig-text-example
#| fig-cap: "Scatterplot showing the size and mass of penguin species in the Palmer Archipelago, Antarctica."
#| warning: false

ggplot(data = penguins,
       aes(x = flipper_length_mm, y = body_mass_g, 
           colour = species)) +
  geom_point() +
  scale_colour_brewer(palette = "Dark2") +
  labs(x = "Flipper length (mm)", y = "Body mass (g)",
       colour = "Species") +
  geom_text(label = "Bold text", y = 6000, x = 180, fontface = "bold",
            colour = "black") +
  geom_label(label = "Italicised \n text", y = 3500, x = 220, 
             fontface = "italic", colour = "black")

```

:::{.callout-warning}
As the `aes` wrapper has been specified at the `ggplot` layer, this will be applied to all `geom`s, including out text and labels. To overwrite this, we can specify the `colour` manually.
:::

:::{.callout-tip}
## Hint

Where text and label positions are determined by data, you may also wish to utilise the `nudge_x` and `nudge_y` arguments to shift annotations and avoid overlap.
:::

Adding text through `geom`s will work but notice that the annotations looks a little blurry on the text. This is because `geom` layers take the data into account and assume that you want the same number of layers/markings as observations in the data. This means that rather than adding a single text or label, `ggplot` is actually adding `{r} filter(penguins, !is.na(body_mass_g), !is.na(flipper_length_mm)) %>% count()`. To overcome this, we can use the `annotate` function instead.

### Annotate function
The `annotate` function will add single `geom` layers to a visualisation while disregarding the rest of the data. This is useful when adding annotations such as text, labels, shapes or arrows. `annotate` functions require the same arguments as the corresponding `geom`, with an additional argument that specifies the `geom` we require. 

For example, we can adapt the code used to create @fig-text-example to utilise the `annotate` functions instead:

```{r}
#| label: fig-text-annotate
#| fig-cap: "Scatterplot showing the size and mass of penguin species in the Palmer Archipelago, Antarctica."
#| warning: false

ggplot(data = penguins,
       aes(x = flipper_length_mm, y = body_mass_g, 
           colour = species)) +
  geom_point() +
  scale_colour_brewer(palette = "Dark2") +
  labs(x = "Flipper length (mm)", y = "Body mass (g)",
       colour = "Species") +
  annotate("text", label = "Bold text", y = 6000, x = 180, 
           fontface = "bold") + #<1>
  annotate("label", label = "Italicised \n text", y = 3500, x = 220, 
             fontface = "italic")

```

1. The `annotate` functions no longer require `colour` to be specified (unless we want to change it) as it is not generated using the data.

Text labels can be combined with curves and arrows to make them clearer, using the `curve` or `segment` `geom`s. Both contain the optional argument `arrow` which adds an arrow to the curved line (this must be defined within the `arrow` function, which can be used to adjust the size or shape of the arrow):


```{r}
#| label: fig-text-curve
#| fig-cap: "Scatterplot showing the size and mass of penguin species in the Palmer Archipelago, Antarctica."
#| warning: false

ggplot(data = penguins,
       aes(x = flipper_length_mm, y = body_mass_g, 
           colour = species)) +
  geom_point() +
  scale_colour_brewer(palette = "Dark2") +
  labs(x = "Flipper length (mm)", y = "Body mass (g)",
       colour = "Species") +
  annotate("curve", x = 185, y = 5500, xend = 195, yend = 4700, 
           curvature = -.3, arrow = arrow(length = unit(.25, "cm"))) +
  annotate("text", x = 185, y = 5600, label = "Important penguin")
```

## Exercise 3 {.unnumbered}
1. Using the `csp_2020` dataset, investigate the relationship between `sfa_2020` and `ct_total_2020`, and show whether that differs between regions. Ensure that this visualisation:

- Has appropriate axes, legend and plot titles
- Has annotations that explain interesting points and make these relationships clearer to a reader

:::{.callout-caution collapse="true"}
## Exercise hint

Include `region` in the `aes` wrapper. Consider adding a line of best fit per region to make these relationships clearer.

This dataset has a clear outlier. Highlight this for readers.

Your `ggplot` should contain `annotate` and `labs` layers.
:::

2. Generate an alternative visualisation that investigates the relationship between `sfa_2020` and `ct_total_2020`, adding an annotation to highlight just the North West region (`region == "NW"`). Add a label to the graph area that makes it clear what the highlighted points represent (rather than a legend).

:::{.callout-caution collapse="true"}
## Exercise hint

Add an extra `geom_point` layer to highlight the North West region.

Use `annotate` layers to define the highlighted point meaning in the graph area. If you are really stuck, check [this ebook](https://ggplot2-book.org/annotations#sec-text-labels) for inspiration.
:::